---
title: "Germination probabilities estimation"
author: "Ingrid Vilmus"
date: "January 9, 2017"
output: html_document
---

Global options

```{r}

```

Package library

```{r}
library(MuMIn)
library(gplots)
library(RColorBrewer)
library(knitr)
library(lattice)
library(lme4)

```

Input File

```{r}
# germination survival
tot<-read.table("/media/rode/SAMSUNG/post_doc Luzerne/donnees/test_germination.csv", header=TRUE, sep=",", dec=".")

# file with informations on the number of seeds per pod for each genotype
pods<-read.table("/media/rode/SAMSUNG/post_doc Luzerne/donnees/Luzerne_graines.csv", header=TRUE, sep=",", dec=".")
pods$Pop_geno_mother<-as.factor(paste(pods$Pop_geno,pods$Mother,sep="_"))
```

Data preparation  
From the greenhouse data, the number of seeds per pod can be averaged by genotype or by mother within genotype (uncomment/comment the corresponding lines to change it).

```{r}

# sum per location --> sometimes, more than one plant germinated from one pod
podsum<-aggregate(tot$GerminationSurvival,by=list(location=tot$location),FUN=sum)
colnames(podsum)[2]<-"GerminationSurvival"
podsum<-merge(tot[,c("ID","zone","quadrat","position","location","Pop_geno")],podsum,all=F)
podsum<-unique(podsum,MARGIN=1)
podsum$Pop_geno_mother<-substr(podsum$ID,start=1,stop=13)

for (i in c(1,6,8)) {podsum[,i]<-as.factor(podsum[,i])}

# average number of seeds per pods per mother 
# seedperpod<-aggregate(pods$Nb_Gr_par_Go,by=list(Pop_geno_mother=pods$Pop_geno_mother),FUN=mean,na.rm=TRUE)
# colnames(seedperpod)[2]<-"seednb"

# or per genotype
seedperpod<-aggregate(pods$Nb_Gr_par_Go,by=list(Pop_geno=pods$Pop_geno),FUN=mean,na.rm=TRUE)
colnames(seedperpod)[2]<-"seednb"


# add to the global data, the germination survival scores
#don_bimod<-merge(podsum,seedperpod,by="Pop_geno_mother",all=FALSE)
don_bimod<-merge(podsum,seedperpod,by="Pop_geno",all=FALSE)

# calculation of the number of death from the initial (estimated) number of seeds per pod and the number of survivors
don_bimod$GerminationDeath<-round(don_bimod$seednb)-don_bimod$GerminationSurvival

# transformation of the quadrat variable since 2 quadrats from 2 different zones can have the same name/nb whereas they should be different
don_bimod$quadrat <- as.factor(paste(don_bimod$zone ,don_bimod$quadrat ,sep="_"))
# delete persisting levels of the factors after subsetting
for (i in c(1:6,8)) {don_bimod[,i]<-factor(don_bimod[,i])}

#write.table(don_bimod,"/media/rode/SAMSUNG/post_doc Luzerne/donnees/pheno_germination.csv",sep=";",dec=".",col.names=T,row.names=FALSE,quote=F)

```


Data exploration  

Visualization of the number of Survival according to the different factors:

- genotypes


```{r}

coul <- c(brewer.pal(9, "Set1"),brewer.pal(nlevels(don_bimod$Pop_geno)-9,"Set3"))
par(mar=c(7.1,4.1,4.1,6.1),xpd=T)
Survmean<-aggregate(don_bimod$GerminationSurvival,by=list(Pop_geno=don_bimod$Pop_geno),mean,na.rm=T)
barplot2(Survmean$x,names.arg = Survmean$Pop_geno,beside=TRUE,col=coul,las=2)
legend("topright",legend = levels(don_bimod$Pop_geno),fill=coul,cex=0.7,inset=c(-0.2,0))

```

The genotypes from Spain seems to have weak germination/survival abilities in the Narbonne context whereas the 3 genotypes from Corsica have the higher number of survivors. However, at this stage, those values are not weighted by the number of seeds present in pods which was demonstrated as being influenced by genotype (H2 individual = 0.67).

- zone and genotypes

```{r}

tabs<-xtabs(don_bimod$GerminationSurvival~ don_bimod$Pop_geno+don_bimod$zone)
coul <- c(brewer.pal(9, "Set1"),brewer.pal(nlevels(don_bimod$Pop_geno)-9,"Set3"))
par(mar=c(5.1,4.1,4.1,6.1),xpd=T)
barplot2(tabs,beside=TRUE,col=coul)
legend("topright",legend = levels(don_bimod$Pop_geno),fill=coul,cex=0.7,inset=c(-0.2,0))

```

Some genotypes were only present in interpop and some only in the zones 12, 13 and 8 (intrapop). The number of survivors is larger in interpop zone since it contains 6 quadrats with 44 locations (1 location corresponds to one pod sown) in each quadrat whereas the zones 12, 3 and 8 contains only 2 quadrats per zone and 32 locations per quadrat.

```{r}
kable(table(don_bimod$zone,don_bimod$quadrat),format="html")
kable(table(don_bimod$zone,don_bimod$Pop_geno),format="html")
```

- quadrat and genotypes


```{r,fig.height=7,fig.width=10}
par(mar=c(5.1,4.1,4.1,2.1),xpd=T)
layout(matrix(c(1,1,2,2),2,2,byrow=T),heights=c(1.5,1))
tabs<-xtabs(don_bimod$GerminationSurvival~ don_bimod$Pop_geno+don_bimod$quadrat)
coul <- c(brewer.pal(9, "Set1"),brewer.pal(nlevels(don_bimod$Pop_geno)-9,"Set3"))

barplot2(tabs,beside=TRUE,col=coul,las=2)
plot.new()
legend("center",legend = levels(don_bimod$Pop_geno),fill=coul,cex=1,inset=c(-0.2,0),ncol=2)


```

At this stage, we can deal with the ratio of Survivors on the number of seeds per pod for each location. First, let's see if there is a link between those variables.

```{r}
Survmean<-merge(seedperpod,Survmean)
par(mar=c(5.1,4.1,4.1,2.1),xpd=T)
layout(matrix(c(1,1,2,2),2,2,byrow=T),heights=c(1.5,1))
plot(Survmean$x~Survmean$seednb,pch=19,xlab="seed nb/pod",ylab="Survivors mean",col=coul)

plot.new()
legend("center",legend = levels(don_bimod$Pop_geno),fill=coul,cex=1,inset=c(-0.2,0),ncol=2)

cortest<-cor.test(Survmean$x,Survmean$seednb)

```

The correlation coefficient is `r round(cor(Survmean$x,Survmean$seednb),2) ` (pvalue = `r round(cortest$p.value,4)`).


We can calculate directly the frequency of germination per genotype using the dataset and considering the number of seeds per pod as a constant per genotype.


```{r}
freq_GS<-aggregate(don_bimod$GerminationSurvival/round(don_bimod$seednb),by=list(Pop_geno=don_bimod$Pop_geno),mean,na.rm=TRUE)
colnames(freq_GS)[2]<-"freq_GS"
par(mar=c(7.1,4.1,4.1,6.1),xpd=T)
barplot2(freq_GS$freq_GS,names.arg = freq_GS$Pop_geno,beside=TRUE,col=coul,las=2,ylab="freq GS")
legend("topright",legend = levels(freq_GS$Pop_geno),fill=coul,cex=0.7,inset=c(-0.2,0))


```


Average per zone and genotype

```{r}
freq_GS_zone<-aggregate(don_bimod$GerminationSurvival/round(don_bimod$seednb),by=list(Pop_geno=don_bimod$Pop_geno,zone=don_bimod$zone),mean,na.rm=TRUE)
colnames(freq_GS_zone)[3]<-"freq_GS"

tabs<-xtabs(freq_GS_zone$freq_GS~ freq_GS_zone$Pop_geno+freq_GS_zone$zone)
coul <- c(brewer.pal(9, "Set1"),brewer.pal(nlevels(don_bimod$Pop_geno)-9,"Set3"))
par(mar=c(5.1,4.1,4.1,6.1),xpd=T)
barplot2(tabs,beside=TRUE,col=coul,ylab="freq GS")
legend("topright",legend = levels(don_bimod$Pop_geno),fill=coul,cex=0.7,inset=c(-0.2,0))


```

Average per quadrat and genotype

```{r,fig.height=7,fig.width=10}
freq_GS_quadrat<-aggregate(don_bimod$GerminationSurvival/round(don_bimod$seednb),by=list(Pop_geno=don_bimod$Pop_geno,quadrat=don_bimod$quadrat),mean,na.rm=TRUE)
colnames(freq_GS_quadrat)[3]<-"freq_GS"

tabs<-xtabs(freq_GS_quadrat$freq_GS~ freq_GS_quadrat$Pop_geno+freq_GS_quadrat$quadrat)
coul <- c(brewer.pal(9, "Set1"),brewer.pal(nlevels(don_bimod$Pop_geno)-9,"Set3"))
layout(matrix(c(1,1,2,2),2,2,byrow=T),heights=c(1.5,1))
par(mar=c(5.1,4.1,4.1,2.1),xpd=T)
barplot2(tabs,beside=TRUE,col=coul,las=2,ylab="freq GS")
plot.new()
legend("center",legend = levels(don_bimod$Pop_geno),fill=coul,cex=1,inset=c(-0.2,0),ncol=2)
```


#### Modelisation of the number of success relative to the number of seeds sown (taken from a previous estimation of the number of seeds per pod for each genotype).  
Germination modelisation with binomial distribution yi -> B(ni,pi) 
yi : number of survivors for the location --> number of success  
pi : probability of survival for the location
ni : average number of grains in a pod per genotype for the location  
ni - yi : number of failures  
The systematic part may contain several variables explanatory variables : zone, quadrat, genotype, mother per genotype.  

1) Choice of the link function on the saturated model based on the AICc

```{r}
m1 <-glm(cbind(GerminationSurvival,GerminationDeath)~quadrat+Pop_geno+zone,data=don_bimod,family=binomial(link='logit'))
m2 <-glm(cbind(GerminationSurvival,GerminationDeath)~quadrat+Pop_geno+zone,data=don_bimod,family=binomial(link='probit'))
m3 <-glm(cbind(GerminationSurvival,GerminationDeath)~quadrat+Pop_geno+zone,data=don_bimod,family=binomial(link='cloglog'))
m4 <-glm(cbind(GerminationSurvival,GerminationDeath)~quadrat+Pop_geno+zone,data=don_bimod,family=binomial(link='cauchit'))

linkfunction<-c('logit','probit','cloglog','cauchit')
names(linkfunction)<-c("m1","m2","m3","m4")

modsel<-model.sel(m1,m2,m3,m4)
modsel

binlink<-linkfunction[rownames(modsel)[1]]

```

2) Choice of the model based on AICc to determine which explanatory variables have to be included  

```{r}
# nul model --> only environmental effect due to quadrat location
germ_bimod0<-glm(cbind(GerminationSurvival,GerminationDeath)~quadrat,data=don_bimod,family=binomial(link=binlink))
summary(germ_bimod0)

# Mother per genotype and quadrat
germ_bimod1<-glm(cbind(GerminationSurvival,GerminationDeath)~quadrat+Pop_geno_mother,data=don_bimod,family=binomial(link=binlink))
summary(germ_bimod1)

# genotype and quadrat
germ_bimod2<-glm(cbind(GerminationSurvival,GerminationDeath)~quadrat+Pop_geno,data=don_bimod,family=binomial(link=binlink))
summary(germ_bimod2)

# genotype and zone
germ_bimod3<-glm(cbind(GerminationSurvival,GerminationDeath)~zone+Pop_geno,data=don_bimod,family=binomial(link=binlink))
summary(germ_bimod3)

# genotype x zone interaction and quadrat
germ_bimod4<-glm(cbind(GerminationSurvival,GerminationDeath)~quadrat+Pop_geno*zone,data=don_bimod,family=binomial(link=binlink))
summary(germ_bimod4)

bimod_sel<-model.sel(germ_bimod0,germ_bimod1,germ_bimod2,germ_bimod3,germ_bimod4)
bimod_sel

bestmod<-get(rownames(bimod_sel[1]))

```

Based on the selection model (AICc criteria), the best model is `r rownames(bimod_sel[1]) ` 


3) Results visualisation and validation  

```{r}
# fitted(bestmod) corresponds to the probabilities predicted by the model for each observation (location)
# comparison of those values with the frequency of germination survival per observation
plot(don_bimod$GerminationSurvival/round(don_bimod$seednb),fitted(bestmod),col=coul[unclass(don_bimod$Pop_geno)],pch=16,xlab="obs",ylab="fitted")

# plot of residuals according to quadrat levels
xyplot(residuals(bestmod)~don_bimod$quadrat)
histogram(~residuals(bestmod)|don_bimod$quadrat)

# according to genotype
histogram(~residuals(bestmod)|don_bimod$Pop_geno)

```

#### Modelisation of the Survival as a Bernouilli variable.  
Germination modelisation with Bernouilli distribution yi -> B(1,pi)  
yi : 1 if the location gave at least one survivor plant and 0 if the location was empty  
pi : probability of survival for the location  
The systematic part may contain several variables explanatory variables : zone, quadrat, genotype, mother per genotype and number of seeds per pod.  


```{r}
don_bimod2<-don_bimod
don_bimod2$GerminationSurvival[don_bimod2$GerminationSurvival==2]<-1

germ_m0<-glm(GerminationSurvival~quadrat,data=don_bimod2,family=binomial)
summary(germ_m0)

germ_m00<-glm(GerminationSurvival~Pop_geno,data=don_bimod2,family=binomial)
summary(germ_m00)

germ_m1<-glm(GerminationSurvival~quadrat+Pop_geno,data=don_bimod2,family=binomial)
summary(germ_m1)


germ_m2<-glm(GerminationSurvival~zone+Pop_geno_mother,data=don_bimod2,family=binomial)
summary(germ_m2)

# add the number of seeds per pods as a covariable
germ_m3<-glm(GerminationSurvival~zone+Pop_geno+seednb,data=don_bimod2,family=binomial)
summary(germ_m3)

germ_msel<-model.sel(germ_m0,germ_m00,germ_m1,germ_m2,germ_m3)
germ_msel

bestmod2<-get(rownames(germ_msel)[1])


```

Data prediction:
if pij predicted <= 0.5, we expected the location is empty and if pij > 0.5, we expected the location contains at least one survivor plant.  

```{r}
pij<-fitted(bestmod2)
pij_label<-as.numeric(pij>0.5)

table(don_bimod2$GerminationSurvival,pij_label)

# wrong classified rate
WC<-sum(pij_label!=don_bimod2$GerminationSurvival)/nrow(don_bimod2)
WC

plot(don_bimod2$GerminationSurvival,fitted(bestmod2),col=coul[unclass(don_bimod$Pop_geno)],pch=16,xlab="obs",ylab="estimated pij")
```

#### Modelisation with random effects 
--> Number of success relative to the number of seeds sown (taken from a previous estimation of the number of seeds per pod for each genotype).  
Germination modelisation with binomial distribution yi -> B(ni,pi) 
yi : number of survivors for the location --> number of success  
pi : probability of survival for the location
ni : average number of grains in a pod per genotype for the location  
ni - yi : number of failures  
The systematic part may contain several explanatory variables : zone, quadrat, genotype, mother per genotype. The random effect can be used for quadrat, genotype and/or mother per genotype effect. Set as random implies we assume the effect follow a Normal distribution with 0 mean and a variance which has to be estimated.  

Choice of the link function  

```{r}
m1 <-glmer(cbind(GerminationSurvival,GerminationDeath)~zone + (1|Pop_geno_mother) + (1|Pop_geno) + (1|zone:Pop_geno)+ (1|quadrat)+ (1|location),data=don_bimod,family=binomial(link='logit'))
m2 <-glmer(cbind(GerminationSurvival,GerminationDeath)~zone + (1|Pop_geno_mother) + (1|Pop_geno) + (1|zone:Pop_geno)+ (1|quadrat)+ (1|location),data=don_bimod,family=binomial(link='probit'))
m3 <-glmer(cbind(GerminationSurvival,GerminationDeath)~zone + (1|Pop_geno_mother) + (1|Pop_geno) + (1|zone:Pop_geno)+ (1|quadrat)+ (1|location),data=don_bimod,family=binomial(link='cloglog'))
m4 <-glmer(cbind(GerminationSurvival,GerminationDeath)~zone + (1|Pop_geno_mother) + (1|Pop_geno) + (1|zone:Pop_geno)+ (1|quadrat)+ (1|location),data=don_bimod,family=binomial(link='cauchit'))

linkfunction<-c('logit','probit','cloglog','cauchit')
names(linkfunction)<-c("m1","m2","m3","m4")

modsel<-model.sel(m1,m2,m3,m4)
modsel

#binlink<-linkfunction[rownames(modsel)[1]]
binlink<-'logit'

```


```{r}

# check for overdispersion
germ_rbimod_od<-glmer(cbind(GerminationSurvival,GerminationDeath) ~ (1|location),data=don_bimod,family=binomial(link=binlink))
summary(germ_rbimod_od)

# environmental effect (quadrat) set as random
germ_rbimod0<-glmer(cbind(GerminationSurvival,GerminationDeath) ~ (1|quadrat),data=don_bimod,family=binomial(link=binlink))
summary(germ_rbimod0)

# genotype effect set as random
germ_rbimod00<-glmer(cbind(GerminationSurvival,GerminationDeath) ~ (1|Pop_geno),data=don_bimod,family=binomial(link=binlink))
summary(germ_rbimod00)

# quadrat and mother per genotype set as random
germ_rbimod1<-glmer(cbind(GerminationSurvival,GerminationDeath) ~ (1|quadrat) + (1|Pop_geno_mother) ,data=don_bimod,family=binomial(link=binlink))
summary(germ_rbimod1)

# quadrat and genotype set as random
germ_rbimod2<-glmer(cbind(GerminationSurvival,GerminationDeath) ~ (1|quadrat) + (1|Pop_geno) ,data=don_bimod,family=binomial(link=binlink))
summary(germ_rbimod2)

# zone and genotype set as random
germ_rbimod3<-glmer(cbind(GerminationSurvival,GerminationDeath) ~ (1|zone) + (1|Pop_geno) ,data=don_bimod,family=binomial(link=binlink))
summary(germ_rbimod3)

# zone as a fixed effect, genotype as a random effect 
germ_rbimod4<-glmer(cbind(GerminationSurvival,GerminationDeath) ~ zone + (1|Pop_geno) ,data=don_bimod,family=binomial(link=binlink))
summary(germ_rbimod4)

relgrad <- with(germ_rbimod4@optinfo$derivs,solve(Hessian,gradient))
max(abs(relgrad))

germ_rbimod4bis<-glmer(cbind(GerminationSurvival,GerminationDeath) ~ zone + (0+zone|Pop_geno) ,data=don_bimod,family=binomial(link=binlink))
relgrad <- with(germ_rbimod4bis@optinfo$derivs,solve(Hessian,gradient))
max(abs(relgrad))

vcov(germ_rbimod4bis)
ranef(germ_rbimod4bis)
summary(germ_rbimod4bis)

# zone as a fixed effect, genotype as a random effect, quadrat and location as random and interaction zone and geno as random
germ_rbimod5<-glmer(cbind(GerminationSurvival,GerminationDeath) ~ zone + (1|Pop_geno) + (1|zone:Pop_geno)+ (1|quadrat)+ (1|location) ,data=don_bimod,family=binomial(link=binlink))
# check for convergence warnings given by glmer, if max(abs(relgrad))<0.01 ok
relgrad <- with(germ_rbimod5@optinfo$derivs,solve(Hessian,gradient))
max(abs(relgrad))



# complete model to estimate variance parameters for simulations
germ_rbimod6<-glmer(cbind(GerminationSurvival,GerminationDeath) ~ zone + (1|Pop_geno_mother) + (1|Pop_geno) + (1|zone:Pop_geno)+ (1|quadrat)+ (1|location) ,data=don_bimod,family=binomial(link=binlink))




# model selection based on aic


germ_msel<-model.sel(germ_rbimod00,germ_rbimod0,germ_rbimod1,germ_rbimod2,germ_rbimod3,germ_rbimod4,germ_rbimod4bis,germ_rbimod5,germ_rbimod6)
germ_msel

bestmod<-get(rownames(germ_msel)[1])
summary(bestmod)

```


Bootstrap procedure to estimate CI

```{r}
?bootMer

mySumm <- function(.) {
   c(beta=fixef(.), sig01=sqrt(unlist(VarCorr(.))))
}

boo01 <- bootMer(germ_rbimod5, mySumm, nsim = 10)

boo01

library(boot)
## Extract all CIs (somewhat awkward)
bCI.tab <- function(b,ind=length(b$t0), type="perc", conf=0.95) {
    btab0 <- t(sapply(as.list(seq(ind)),
                    function(i)
                    boot.ci(b,index=i,conf=conf, type=type)$percent))
    btab <- btab0[,4:5]
    rownames(btab) <- names(b$t0)
    a <- (1 - conf)/2
    a <- c(a, 1 - a)
    pct <- stats:::format.perc(a, 3)
    colnames(btab) <- pct
    return(btab)
}
bCI.tab(boo01)
```


